---
title: "Data Structures"
author: "Nicholas Ho, Richard Morris, Darya Vanichkina"
keypoints:
- R supports multiple variable types
- Errors often result because of trying to perform an unsupported operation on a specific
  data type
- Errors can be cryptic to interpret
- We can use helper packages to import and filter data in R
objectives:
- Assign values to objects in R.
- Solve simple arithmetic operations in R.
- Describe what a data frame is.
questions: Working with more realistic data
source: Rmd
teaching: 50
exercises: 25
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

## Working with more realistic Data

The dataset that we will be using today is taken from University of California, Irvine's Machine Learning Repository. We will be examining the [Autism Screening Adult Data Set](https://archive.ics.uci.edu/ml/datasets/Autism+Screening+Adult). This is data collected by Dr Fadi Fayez Thabtah at NMIT via the mobile app ASD Tests.

Click [here](https://raw.githubusercontent.com/Sydney-Informatics-Hub/lessonbmc/gh-pages/_episodes_rmd/data/autism_data.csv) to open the data on your browser. Then, right-click and Save As "autism_data" in the "comma-separated values" or "csv" format into a data/ folder in your working directory.

Variable | Definition
-----------------|------------------------------------------------
id | The participant's ID number
AX_Score | Score based on the Autism Spectrum Quotient (AQ) 10 item screening tool [AQ-10](http://docs.autismresearchcentre.com/tests/AQ10.pdf)
age | Age in years
gender | Male or Female
ethnicity | Ethnicities in text form
jaundice | Whether or not the participant was born with jaundice?
autism | Whether or not anyone in tbe immediate family has been diagnosed with autism?
country_of_res| Countries in text format
used_app_before | Whether the participant has used a screening app
result | Score from the AQ-10 screening tool
age_desc | Age as categorical
relation | Relation of person who completed the test
Class/ASD | Participant classification


### Reading in Data

The first thing to do before we can begin an analysis is loading some data. To do this we can use the below command.

```{r read_data_1}
autism.data <- read.csv(file = "data/autism_data.csv",
                        header = TRUE,
                        sep = ",",
                        na.strings = "?")
```

Before going any further, let's break down what this command does specifically.

1. It reads the file, which could be
- a file path e.g. `"~/Documents/MyFolder/datafile.csv"`
- a url like we have here e.g. `"http://www.website.com/subdomain/datafile.csv"`
2. It specifies that the first row in the csv file contains "header" information i.e. column names
3. It specifies that neighbouring columns are separated by a comma ","
4. It specifies that this CSV file uses "?" as the text for missing data

How would we find this out if we didn't know already? Look at the help file:
```{r eval=FALSE}
?read.csv
```

Comma Separated Value (.csv) files are usually the standard, simplest format which is compatible with all sort of different software, e.g. R, python, Excel, MATLAB, ...

But if we needed to read in a different format of data, there's usually a function or a library for doing just that,

e.g.
in base R:

- `read.csv()`
- `read.table()`

In the `xlsx` library:

- `read.xlsx()`

In the `haven` library:

- `read_spss()` for SPSS files

In the `tidyverse` family of libraries (more specifically, part of the `readr` library):
- `read_csv()`
- `read_tsv()`
- `read_delim()`
- `read_fwf()`
- `read_table()`


### Viewing The Data
Remember, we use the function `View()` to visually inspect at the data
```{r, eval = FALSE}
View(autism.data)
```

If the dataset is large, it is usually more prudent to view only the first few rows, using the head command:


```{r}
head(autism.data)
```

What data type is this data?
```{r}
class(autism.data)
```


What variables are included in this dataset?
```{r}
colnames(autism.data)
```


### Changing The Column Names
We've a typo in the country of residence column. Let's correct this by first finding the column with the typo.
```{r}
colnames(autism.data)

which(colnames(autism.data) == "contry_of_res")

colnames(autism.data)[17] <- "country"
```

Afterwards, view the data again to check the column names have been changed
```{r eval=FALSE}
View(autism.data)
```

### Add a new column
Let's create a better participant ID than just 1,2,3,...

The first function we will need is `paste()`. This function concatenates strings. For example;
```{r}
paste("hel", "lo")
```


This leaves an unwanted gap, hence we'll use the `sep` parameter
```{r}
paste("hel", "lo", sep = "")
```

A shorthand for this command is called paste0, but we're showing the `paste()` command here, since you may want to use other delimeters, such as "@", for example, in your own work.

Now let's add the letters "PatientID_" to the front of the IDS. Then, we'll add this as a new column

```{r}
autism.data$pids <- paste( "PatientID_" , autism.data$id, sep = "")
```

Afterwards, view the data again to check the new "pids" column.
```{r eval=FALSE}
View(autism.data)
```



### Accessing Subsets
Let's look at all the records for those with a family history of autism:
```{r}
family.autism.data <- autism.data[autism.data$autism == "yes",]
```

What is the average AQ-10 score for those with a familial history of autism?
```{r}
mean(family.autism.data$result)
```

What about for those with no familial history of autism?
```{r}
no.family.autism.data <- autism.data[autism.data$autism == "no",]
mean(no.family.autism.data$result)
```

Who is the oldest person in the dataset and what is their age?
```{r}
max(autism.data$age)

# ignore missing data
max(autism.data$age, na.rm = TRUE)
which.max(autism.data$age)
```

Someone is apparently 383 years old! There are a number of possibilities why this value is here with a range of probabilities from the most likely (data entry error) to the least (the participant is yoda).

For our purposes, let's remove this person. To do so, we'll use the `-` sign in front of the indices we want to remove.
```{r}
autism.data <- autism.data[-which.max(autism.data$age), ]
```

Who is the youngest person in this dataset and how old are they?
```{r}
min(autism.data$age, na.rm = TRUE)

which.min(autism.data$age)
```


### Summarising Your Data
A handy way to inspect your data is to summarise it. To do so, use the `summary()` function
```{r}
summary(autism.data)
```


Use the summary function to return a quick summary of only the country column
```{r}
summary(autism.data$country)
```


Why does the summary function give you quartiles for some column and provide counts for others? Remember that data frames are 2D collections that can hold data of different types
```{r}
class(autism.data$age)
summary(autism.data$age)


class(autism.data$country)
summary(autism.data$country)
```


Another way to summarise a `factor` column is to use the `table()` function
```{r}
table(autism.data$country)
```


A difference between using the `summary()` function over the `table()` function for factors is that by default, the `summary()` function also counts `NA` values whereas in this must be specified for `table()`
```{r}
summary(autism.data$ethnicity)

table(autism.data$ethnicity)

table(autism.data$ethnicity, useNA = "ifany")
```


###Quiz




> ## Section quiz
>
> 1. What is the proportion of males to females in this dataset?
>
> 2. Return those participants whose `gender` is female
>
> 3. Return those participants whose `gender` is female into a new data frame called "females.autism.data"
>
> 4. What is the mean `result` score for those classified as ASD
>
> 5. What is the mean `result` score for those not classified as ASD
>
> {: .source}
>
> > ## Solution
> >
> > 1.  What is the proportion of males to females in this dataset?
> >
> > ~~~
> > table(autism.data$gender)
> > prop.table(table(autism.data$gender))
> > ~~~
> > 2. Return those participants whose 'gender' is female
> >
> > ~~~
> > autism.data[autism.data$gender == "f", ]
> > ~~~
> > 3. Return those participants whose 'gender' is female into a new data frame called "females.autism.data"
> >
> > ~~~
> > females.autism.data <- autism.data[autism.data$gender == "f", ]
> > ~~~
> > 4.  What is the mean 'result' score for those classified as ASD
> >
> > ~~~
> > mean(autism.data[autism.data$Class.ASD == "YES" ,"result"])
> > ~~~
> > 5. What is the mean 'result' score for those not classified as ASD
> > 
> > ~~~
> > mean(autism.data[autism.data$Class.ASD == "NO" ,"result"])
> > ~~~
> > {: .output}
> {: .solution}
{: .challenge}




## Writing the data out
Use the write.table() function to write the data into a text file

```{r}
write.table(x = autism.data,
            file = "data/autism_pids.txt",
            row.names = FALSE)
```

Use the write.csv() function to make the datafile a csv (comma separated values) file
```{r}
write.csv(x = autism.data,
          file = "data/autism_pids.csv",
          row.names = FALSE)
```
