---
title: "Data and statistical analysis"
author: "Nicholas Ho, Darya Vanichkina, Richard Morris, Maryam Montazerolghaem"
keypoints:
- R has a large community of developers creating power tools for data and statistical analysis
objectives: 
- Introduction into using R for statistical analysis and visualisation
questions: Using R for data and statistical analysis
source: Rmd
subtitle: Day 3
teaching: 60
exercises: 25
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("10-")
```


## Data and statistical analysis

We have previously shown you how to wrangle your data into the right shape with the tidyverse. Now onto hypothesis testing and statistical analysis. 


Our aim here is not to teach you statistics, but how to use R to perform the most popular statistical analysis on your data. For statistical consultation, please contact Sydney Informatics Hub's statistics team. 

"To consult the statistician after an experiment is finished is often merely to ask him to conduct a post mortem examination. He can perhaps say what the experiment died of"
  - Ronald Fisher 1938


The data for this session can be [downloaded from here](https://raw.githubusercontent.com/Sydney-Informatics-Hub/lessonbmc/gh-pages/_episodes_rmd/data/gait_clean.csv). This is a selection of datapoints from a database of demographic and clinical measurements of Parkinson's patients and controls. More information about this dataset can be found at http://physionet.org/physiobank/database/gaitpdb/.

Please download this .csv file into your working directory's `data` folder.

```{r, warning=FALSE, message = FALSE}
library(tidyverse)

gait <- read.csv("data/gait_clean.csv", 
                 header = TRUE,
                 row.names = 1)

# another way to read in this CSV file is to use the URL option in read.csv()
gait <- read.csv("https://raw.githubusercontent.com/Sydney-Informatics-Hub/lessonbmc/gh-pages/_episodes_rmd/data/gait_clean.csv", 
                 header = TRUE,
                 row.names = 1)

# let's inspect the data
summary(gait)
```


## Correlation

We're interested to see if there's any correlation between two clinical measures: UPDRS (Unified Parkinson's Disease Rating Scale) and TUAG (Timed Up And Go Test)

```{r, warning=FALSE, message = FALSE}
cor(gait$UPDRS, gait$TUAG, method = "spearman")

ggplot(gait, aes(x = TUAG, y = UPDRS)) +
  geom_point()
```

> ## Section quiz
> 1. What's the Spearman's rank correlation coefficient between UPDRS and UPDRSM? 
> {: .source}
> > ## Solution
> > 1. 
> > ~~~
> > cor(gait$UPDRS, gait$UPDRSM, method = "spearman")
> > ~~~
> > {: .output}
> {: .solution}
{: .challenge}



We're interested to see if there's any correlation between any of the clinical variables in the dataset. Let's explore this by generating a correlation plot.
```{r, warning=FALSE, message = FALSE}
# please install corrplot prior to running the below code by running: install.packages("corrplot")
library(corrplot)

# keep the clinical variables
gait_clin <- gait %>%
  select(HoehnYahr, UPDRS, UPDRSM, TUAG)

# create a correlation matrix of the clinical markers using Spearman correlation
correlations <- cor(gait_clin, method = "spearman")

# create the correlation plot
corrplot(correlations, method = "color")
```


## Hypothesis testing 


Let's say we want to simply compare the `TUAG` score between `Group` by using the 2-sample t-test. Our null hypothesis is that there is no difference between the mean `TUAG` of the Parkinson's and Control participants. By default in R, the `t.test()` function will perform Welch's 2-sample t-test. We can plot this with a violin plot
```{r, warning=FALSE, message = FALSE}
t.test(TUAG ~ Group, data = gait)
  
# Let's visualise this data
ggplot(gait, aes(y=TUAG, x = Group)) + 
  geom_violin() 

# have a bit of fun and change the plot to the Wall St Journal's style
library(ggthemes)
ggplot(gait, aes(y=TUAG, x = Group,fill = Group)) + 
  geom_violin() + 
  theme_wsj()

```
> ## Section quiz
> 1. Perform the Welch 2-sample t-test for TUAG by gender? Generate violin plots of this.
> {: .source}
> > ## Solution
> > 1. 
> > ~~~
> > t.test(TUAG ~ Gender, data = gait)
> > ggplot(gait, aes(y=TUAG, x = Group)) + geom_violin() 
> > ~~~
> > {: .output}
> {: .solution}
{: .challenge}


We can perform the non-parametric Wilcoxon-Mann-Whitney Test. Here the null hypothesis is that the distributions of the two groups are the same.
```{r, warning=FALSE, message = FALSE}
wilcox.test(TUAG ~ Group, data = gait)
```


We can then perform an ANOVA using the `Study` grouping of patients. Here, we are testing the null hypothesis that the mean `TUAG` is the same for all groups.
```{r, warning=FALSE, message = FALSE}
aov_study <- aov(TUAG ~ Study, data = gait)
summary(aov_study)

ggplot(gait, aes(y=TUAG, x = Study)) + 
  geom_violin() 
```


Let's assume that the we had sufficient evidence from the ANOVA to reject the null. We can then perform post-hoc tests. For example, we can perform multiple pairwise-comparison between the means of groups using Tukey Honest Significant Differences
```{r, warning=FALSE, message = FALSE}
TukeyHSD(aov_study)
```


R has a variety of statistical tests available. Google is your best friend here.


## Linear regression

Let's say that for the Parksinson's patients, we want to make a model to predict `UPDRS` score using the `UPDRSM` 
```{r, message = FALSE, warning = FALSE}
gait_pd <- gait %>%
  filter(Group == "PD")

# Let's plot this first
ggplot(gait_pd, aes(x = UPDRSM, y = UPDRS)) +
  geom_point() +
  geom_smooth(method = 'lm', se = TRUE)
```

Fit a linear regression predicting UPDRS using UPDRSM, and look at the summary of the results
```{r, warning=FALSE, fig.show = 'hide', message = FALSE}
gait_pd_slr <- lm(UPDRS ~ UPDRSM, data = gait_pd)
summary(gait_pd_slr)

gait_pd_slr$coefficients

plot(gait_pd_slr)
```


What if we now include `Age` in our model
```{r, warning=FALSE, fig.show = 'hide', message = FALSE}
gait_pd_mlr <- lm(UPDRS ~ UPDRSM + Age, data = gait_pd)
summary(gait_pd_mlr)

gait_pd_mlr$coefficients

plot(gait_pd_mlr)
```


## Visualisation with PCA

We will use PCA and plot the first 2 Principal Components to observe the variance in a microarray dataset. This is a famous dataset from 1999 where 27 acute lymphoblatic leukaemia (ALL) and 11 acute myeloid leukaemia (AML) patient's gene expression data were measured for 3051 genes.

First, we need to install and load up the `multtest` package which comes with the `golub` data.
```{r, message = FALSE}
# please install multtest prior to running the below code by running: install.packages("multtest")
library(multtest)
data(golub)

# this is a vector of patient's diagnoses. ALL is 0 and AML is 1
table(golub.cl)

# let's try PCA in R for visualisation. First, we have to transpose the dataset because we want each patient to be recorded in rows and genes in columns
golub <- t(golub)

# perform PCA with the input data z-scored such that the mean for each gene is 0 and the SD for each gene is 1
pca <- prcomp(golub, scale = TRUE, center = TRUE)

# I'm just interested in plotting the Principal Components. This is in the "x" matrix of the pca object
pca_plotdata <- data.frame(pca$x)

# attach the patient diagnosis as we will be colouring in our plot with this
pca_plotdata <- pca_plotdata %>%
  mutate(samples = as.factor(golub.cl)) 

# plot PC1 and PC2 with ggplot
ggplot(pca_plotdata, aes(x = PC1, y = PC2, colour = samples)) +
  geom_point()

# inspect the cumulative proportion of variance explained by each PC
summary(pca)

# quickly plot the cumulative proportion of variance explained using base R's plotting function
plot(summary(pca)$importance["Cumulative Proportion",])
```


## Clustering and heatmap
```{r, message = FALSE}
# please install gplots prior to running the below code by running: install.packages("gplots")
library(gplots)

# my favourite genes are in these columns
favgenes <- c(703,717,766,829,896,1037,1334,1665,1817,1834,2002,2124,2386,2600,2645,2801,2851,2939)

# let's make a heatmap of my favourite genes
heatmap.2(golub[ ,favgenes], 
          trace = "none",
          scale = "column",
          RowSideColors = c(rep("yellow", 27), rep("darkgreen", 11)),
          dendrogram = "both", 
          col = "bluered")
```



